---
title: "Multi-Scale Optimal Control Update"
author: "Noam Ross"
date: "7/5/2015"
output: html_document
---

# 1. Boundary values and mutiple solutions to the ODE Hamiltonian

***Is this optimal path feasible or due to a numerical error***

The comparison case for my stochastic, individual-based optimal control problem
has been this ODE-based optimal control problem.  

-  System dynamics ($N$ = hosts, $P$ = parasites):

$$\begin{align}
\frac{dN}{dt} &= r N (1 - N/K) - \alpha P - d N \\
\frac{dP}{dt} &= \lambda P N - \mu P - d P - \alpha P - \alpha P^2/N + e^{-h} N \lambda_{ex}
\end{align}$$ 


-  Profit function (continuous provision of ecosystem pervice per individual host minus control cost):

$$vN - ch$$

-  Hamiltonian


$$\begin{align}
\mathcal{H} = &vN - ch + \\
&\nu_1 \left( rN(1 - N/K) - \alpha P - d N \right) + \\
&\nu_2 \left(\lambda P N - \mu P - d P - \alpha P - \alpha (P^2)/N + e^{-h}  N \lambda_{ex} \right)
\end{align}$$

-  Shadow price dynamics (via adjoint principle):

$$\begin{align}
\frac{d\nu_1}{dt} &= -v - \nu_1 \left(r - d - 2r \frac{N}{K}\right) - \nu_2 \left(\lambda P + \alpha \frac{P^2}{N^2} + e^{-h} \lambda_{ex} \right) + \delta \nu_1 \\
\frac{d\nu_2}{dt} &= \nu_1 \alpha - \nu_2 \left(\lambda N - \mu - d - \alpha - \frac{2 \alpha P}{N} \right) + \delta \nu_2
\end{align}$$

-  The control, which reduces arrival rate of new parasites or spores, is determined
by solving

$$ h \frac{d\mathcal H}{dh} = 0 = h \left( -c - \nu_2 e^{-h} N * \lambda_{ex} \right)$$

Since the problem is solved over a fixed time period, and there are no policy
constraints on the terminal conditions, I don't have boundary conditions for the
state or shadow variables, so I solve for initial values of $\nu_1$ and $\nu_2$ by maxmizing
the overall profit, calculated as

$$\int_{t=0}^{100} vN(t) - ch(t) dt$$

I calculate this integral using cubic splines over the grid of the solution.

```{r setup, echo=FALSE, include=FALSE}
#devtools::install_github('noamross/tracer')
#devtools::install_github('noamross/noamtools')
library(deSolve)
library(bvpSolve)
library(rootSolve)
library(nloptr)
library(tracer)
library(ggplot2)
library(noamtools)
library(dplyr)
library(tidyr)
knitr::opts_chunk$set(cache=TRUE)
knitr::opts_chunk$set(warning = FALSE)
```

```{r def-eqns, echo=FALSE}
syseq2 = function(t, state, parms) {
  with(as.list(c(state, parms)), {
    h_calc = c(parms$control_min, parms$control_max, uniroot.all(hH_h,c(parms$control_min, parms$control_max), n=10000, t=t, state=state, parms=parms))
    if(length(h_calc) == 3) h_calc = c(h_calc, NA)
    H_calc = H(h_calc, t, state, parms)
    if(length(H_calc) == 3) H_calc = c(h_calc, NA)
    H_sel = max(H_calc, na.rm=TRUE)
    h = h_calc[which.max(H_calc)]
    dN = max(r*N*(1 - N/K), 0) - alpha*P - d*N
    if(isTRUE(all.equal(dN, 0))) dN = 0
    dP = lambda*P*N - mu*P - d*P - alpha*P - alpha*(P^2)/N + exp(-h)*N*lambda_ex
    dS1 = -v - S1*(r - d - 2*(r/K)*N) - S2*(lambda*P + alpha*(P^2/N^2) + exp(-h) * lambda_ex) + delta*S1
    dS2 = S1*alpha - S2*(lambda*N - mu - d - alpha - 2*alpha*P/N) + delta*S2
    derivs = c(dN=dN, dP=dP, dS1=dS1, dS2=dS2)
    #    derivs = ifelse(abs(derivs) > .Machine$double.eps ^ 0.5, derivs, 0)
    return(list(derivs, c(derivs, h_calc = h_calc, H_calc = H_calc, H_sel=H_sel, h=h)))
  })
}

hH_h = function(h, t, state, parms) {
  with(as.list(c(state, parms)), {
    h*( - cost - S2 * exp(-h) * N * lambda_ex)
  })
}

H = function(h, t, state, parms) {
  with(as.list(c(state, parms)), {
    val = v*N - cost*h + S1*(r*N*(1 - N/K) - alpha*P - d*N) + S2*(lambda*P*N - mu*P - d*P - alpha*P - alpha*(P^2)/N + exp(-h)*N*lambda_ex)
    return(val)
  })
}
```

```{r setparms, echo=FALSE}

parms = list(
  lambda = 0.001,
  lambda_ex = 0.2,
  alpha = 0.1,
  mu = 0.01,
  r = 0.5,
  d = 0.01,
  K = 100,
  control_min = 0,
  control_max = 1000,
  v = 50,
  cost = 200,
  c = 200,
  delta = 0,
  progress = TRUE
  #micro_record = file("micro.txt", open="w+")
  #  macro_record = file("macro.txt", open="w")
)


N0 = 100
P0 = 0
S10 =  95.325211
S20 =  -465.441918
init = c(N = N0, P = P0, S1 = S10, S2 = S20)
times = seq(0,100, by=1)
```

```{r optfuns, echo=FALSE}
profout = function(out) {
  intval = integrate(splinefun(out[,"time"], (out[,"N"]*parms$v - (out[,"h"])*parms$cost)*exp(-out[,"time"]*parms$delta)),
            lower = out[1,"time"], upper=tail(out[,"time"], 1))
  return(intval$value)
}

profun = function(ivals, sys, parms) {
  init = c(N=N0, P=P0, S1=ivals[1], S2=ivals[2])
  out = lsoda(init, times, sys, parms)
  -profout(out)
}

shadzerofun = function(ivals, sys, parms) {
  init = c(N=N0, P=P0, S1=ivals[1], S2=ivals[2])
  out = lsoda(init, times, sys, parms)
  
}
```

This optimization successfully converges:

```{r optimize, echo = FALSE}
library(tracer)
# #
opt0 <- nloptr_tr(x0 = c(S10,S20), eval_f = profun, lb=c(-1e7, -1e7), ub=c(1e7, 1e7),
               opts = list(algorithm = "NLOPT_GN_CRS2_LM", maxeval=300), sys=syseq2, parms=parms)
opt0trace = tracer(opt0)
opt = nloptr_tr(x0 = opt0$solution, eval_f = profun, lb=c(-Inf, -Inf), ub=c(Inf, Inf),
             opts = list(algorithm = "NLOPT_LN_SBPLX", maxeval=1000), sys=syseq2, parms=parms)

opttrace = tracer(opt)
opt
```

Above, the values `r round(opt$solution[1], 3)` and `r round(opt$solution[2], 3)` are the initial shadow values. 

This yields the following optimal path:

```{r baserun, echo = FALSE}

out_opt = lsoda(c(N = N0, P = P0, S1 = opt$solution[1], S2 = opt$solution[2]), times, syseq2, parms)
#matplot(out_opt[,1], out_opt[,c("N", "P", "h")], type="l", lty=1, lwd=1, main=round(profout(out_opt)))

out = as.data.frame(out_opt)
out_tidy = gather(out, "variable", "value", -time)

```

```{r plot-path, echo = FALSE}
ggplot(filter(out_tidy, variable %in% c("N", "P", "h")),
       aes(x=time, y=value, col=variable)) + 
  geom_line() +
  labs(list(x = "Time", y = "Population or Control level")) +
  theme_nr
```

I sanity-checked these by using a shooting algorithm to finish the simulation at similar (not identical) finish points, which yields the same path.

Here is the path of the shadow values, which I show at 3 scales to show how they change
and then go very negative very fast.  Note that when the shadow values for the 
parasite go *positive* is when the optimal control falls to zero.  This positive value is counter-intuitive but
makes some sense.  At that point in the epidemic, additional parasites will make
the epidemic burn out quicker.

```{r plot-shadow, echo = FALSE}
ggplot(filter(out_tidy, variable %in% c("S1", "S2")),
       aes(x=time, y=value, col=variable)) + 
  geom_line() +
  labs(list(x = "Time", y = "Shadow value")) +
  theme_nr + ylim(-2e3, 2e3)

ggplot(filter(out_tidy, variable %in% c("S1", "S2")),
       aes(x=time, y=value, col=variable)) + 
  geom_line() +
  labs(list(x = "Time", y = "Shadow value")) +
  theme_nr + ylim(-2e3, 3e5)


ggplot(filter(out_tidy, variable %in% c("S1", "S2")),
       aes(x=time, y=value, col=variable)) + 
  geom_line() +
  labs(list(x = "Time", y = "Shadow value")) +
  theme_nr

```

Note that the shadow value of the parasites goes *positive* at this point, which
is counter-intuitive but I think makes sense.

Since the solution to $h \, d\mathcal{H}/ dh$ has multiple roots, I solve for both
and determine the value of $h$ which gives the highest value for $\mathcal{H}$.  Here is a 
plot of the $\mathcal{H}$ over time, as well as the control level:

```{r plot-hamil, echo = FALSE}
ggplot(filter(out_tidy, variable %in% c("H_calc3", "H_calc4")),
       aes(x=time, y=value, col=variable)) + 
  geom_line() +
  labs(list(x = "Time", y = "Hamiltonian")) +
  theme_nr + ylim(-20000, 5000)
```


```{r plot-control, echo = FALSE}
ggplot(filter(out_tidy, variable %in% c("h_calc3", "h_calc4")),
       aes(x=time, y=value, col=variable)) + 
  geom_line() +
  labs(list(x = "Time", y = "Control Level")) +
  theme_nr
```

# 2. Numerical stability in calculating shadow values in the IBM

I've been working on improving the numerical stability of my IBM-optimization
algorithm.  Up until now, this has been a matter of speeding up code so as
to reduce the grid size and increase the number of simulations I use to 
calculate the derivatives of the mean IBM behavior.  However, I encountered a 
challenge in calcualting shadow values for which this doesn't help.

Shadow values are calculated via the adjoint equation:

$$\frac{d\nu}{dt} = - \frac{d\mathcal{H}}{dN} = - \frac{d\pi}{dN} - \nu \frac{dN'}{dN}$$

I approximate \frac{dx'}{dx} as

$$ \frac{dN'}{dN} = \frac{dN^2 / d^2t}{dN/dt} \approx \frac{\Delta N'}{\Delta N}$$

However, when the system has more than one state variable, this approximation
doesn't hold because $\frac{dN'}{dN}$ needs to be calculated for each variable
while holding the other constant, which isn't possible when calculating only
via forward simulation.

However, I'm currently implementing another approach that takes advantage of the fact
that the macro-level system representation is of two continuous variables ($N,P$),
but the micro-level simulation is of discrete counts.

The "lifting" function generates a number of parallel instances of the micro-level
simulation with discrete counts from the continuous macro-level variables.  If
the macro-level population in 18.7, it will generate 30% of the simulations starting
with a population of 18 and 70% with a population, and average the results of all of
them.  This means I can directly calculate $\frac{dN'}{dN} \approx \frac{\Delta N'}{\Delta N}$
using the difference in $N$ between the different simulations.  Over a range <1, this will
be linear.  In order to calculate with accuracy without increasing the number of simultions,
I can simply allocate half the simulations to each discrete level, and use a weighted average
to calculate outcomes.

```{r optshoot, echo=FALSE}
shootfun = function(ivals, sys, target, parms) {
  init = c(N=N0, P=P0, S1=ivals[1], S2=ivals[2])
  out = lsoda(init, times, sys, parms)
  obj = sum((tail(out, 1)[,c("S1", "S2")] - target)^2)
  if(any(is.na(obj))) obj = Inf
  return(obj)
}

opt0_s <- nloptr_tr(x0 = c(S10,S20), eval_f = shootfun, lb=c(-1e7, -1e7), ub=c(1e7, 1e7),
               opts = list(algorithm = "NLOPT_GN_CRS2_LM", maxeval=300), sys=syseq2, parms=parms, target = c(0,0))
opt0_s_trace = tracer(opt0_s)
opt_s = nloptr_tr(x0 = opt0_s$solution, eval_f = shootfun, lb=c(-Inf, -Inf), ub=c(Inf, Inf),
             opts = list(algorithm = "NLOPT_LN_SBPLX", maxeval=1000), sys=syseq2, parms=parms, target = c(0,0))

opttrace = tracer(opt_s)
opt_s

```
