---
title: "Optimal Control of Individual-Based Disease Models"
author:
- affiliation: UCD
  email: nmross\@ucdavis.edu
  footnote: Corresponding author
  name: Noam Ross
date: "August 17, 2015"
output:
  pdf_document:
    dev:
    - pdf
    fig_caption:
    - yes
    fig_height:
    - 6
    fig_width:
    - 11
    keep_tex:
    - yes
    template:
    - elsarticle.latex
csl: ecology-letters.csl
address:
- address: |
    Department of Environmental Science and Policy
    University of California-Davis
    1 Shields Avenue Davis, CA 95616 USA
  code: UCD
documentclass: elsarticle
layout: review,12pt
linenumbers: yes
bibliography: references.bib
abstract: "Optimal control techniques can be used to derive cost-effective management
  strategies for wildlife disease, but require appropriate models for disease dynamics.
  These dynamics can be driven by driven by mechanisms such as population structure,
  individual variation, and stochasticity, which can be difficult to represent in
  closed-form equations and are often best simulated using individual-based models
  (IBMs), whose use has been expanding in both applied and theoretical ecology.  Yet
  applying optimal control to IBMs is challenging due to IBMs' high dimensionality,
  which precludes robust exploration of state space. \"Equation-free\" (EF) methods
  can be used to reduce IBM states to a small number of variables even when IBMs are
  only represented by simulation code. Here, I demonstrate a new use of EF methods
  in determining optimal control of IBMs. I apply these methods a management problem
  of conserving a host population where individuals vary in disease load and are subject
  to exogenous force of infection, which may be reduced by management effort. I show
  that the EF method can recover the same optimal control strategy via analytical
  methods without the need to derive closed-form equations for population dynamics."
---


```{r get_chunks, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE}
library(knitr)
read_chunk('base_parms.R')
read_chunk('compare_graphs.R')
read_chunk('ode_soln.R')
opts_knit$set(eval.after = "fig.cap")
opts_chunk$set(echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE)
```

# Introduction

A central challenge in disease ecology is developing effective management 
strategies to control harmful diseases and limit their damage to wildlife, 
human health, and ecosystem services [@Joseph2013].   Disease management may be considered an economic problem: Given limited resources, what is the best strategy to minimize minimize impacts? Or, conversely, given management targets, how to best limit the cost of disease control? Research addressing such questions - economic epidemiology - is a growing field [@Brandeau2003; @Rowthorn2009], though the literature to on the economics of wildlife disease is more limited] [@Bicknell1999; @Horan2005; Fenichel2010; @Horie2013]. The bioeconomics of invasive species management is more broadly explored [@Sharov1998; @Carrasco2010; @Haight2010; @EpanchinNiell2012], and has many close parallels with the economics of disease management.

Individual-based models (IBMs, also called agent-based models) simulate systems 
based on the behavior and properties of individuals, rather than properties at the population scale. IBMs can simulate system behavior arising from individual behaviors and interactions difficult to describe in aggregate, including network contact structure, spatial heterogeneity, and variation in individual traits, all of which can affect disease dynamics. These individual behaviors and interactions 
result in emergent dynamics at the population scale that often can not be 
expressed as closed-form equations at without making strong assumptions about the distribution of individuals [@Grimm2005].

IBMs can and have been used to study disease economics in a number of ways.  Most commonly, is to simulate IBMs of disease systems under a 
number of discrete management scenarios and examine the outcomes of each 
scenario to determine which does best according to performance criteria [@Okell2008; @Mao2011; @Yang2011]. IBMs can also incorporate adaptive or optimizing behavior within individuals in order to better simulate disease dynamics [@Fenichel2010]. 

Determining optimal control approaches using IBMs is more challenging. IBMs also suffer from "the curse of dimensionality": high dimensionality results in very large state space that can not be explored thoroughly. In the absence of closed-form equations of overall system behavior, one can not derive optimal control decisions analytically. 
 
One approach to optimal controlling using IBMs has been to apply
analytical control methods to reduced-form models that approximate IBM behavior. The optimization problem can be solved on a this set of ordinary differential equations that represent an approximation of the IBM. The efficacy of this approach depends on how well reduced-form models approximate IBM behavior.  Mean-field approximations may work when IBMs represent well-mixed or homogenous systems, but perform poorly under heterogeneity [@Federico2013].  Another approach is to generate reduced-form models that use non-mechanistic equations to more closely approximate IBM behavior. Such methods can yield better results than mean-field models. However, the applicability of these adjusted models is limited to the range over which they were parameterized, and model parameters are difficult to interpret [@Oremland2015].

Here I describe another approach to deriving optimal control strategies for 
ecological systems using IBMs that uses equation-free, or multi-scale, modeling 
[@Armaou2004] to estimate mean dynamics of an IBM numerically and estimate 
quantities that allow the use of analytic tools to determine an optimal control 
path. First, I introduce the equation-free modeling framework. I then 
describe a disease system and equivalent ordinary differential equation (ODE) and IBM models. I solve the ODE system using classic tools and the IBM model using the methods described, showing that they generate the same solution.

## Equation-free modeling

Equation-free (EF), or multi-scale, modeling [@Armaou2004; but see @Kevrekidis2009a for review] is a method for capturing and analyzing  population-scale dynamics of IBMs while bypassing the deriving population-scale equations. It has been used in a variety of biological applications, including evolution, movement ecology, and epidemiology [@Cisternas2004; @Erban2006; Gross2008a; Raghib2010; @Williams2015]. In EF modeling, the population-scale dynamics are estimated using short, repeated bursts of the IBM, which are used to estimate population-scale derivatives of the system. 

For each IBM burst, a "lifting" operator maps the population-scale state to 
a distribution of possible individual-scale states. Each of these instances of the IBM is simulated for a short period of time, then all IBM instances' updated states are converted back to population-scale states and aggregated with a "restricting" function. Information about the IBM state is lost in restriction and generated by assumptions in the lifting function. Thus, EF is another form of model reduction, albeit one that offers alternatives to mean-field and other approximations. Lifting and restriction operators may be chosen to capture the important heterogeneities represented in the IBM and relevant to the problem at hand. While these functions must be chosen with care, there is greater latitude in the assumptions they contain than would be needed to reduce models analytically. In addition, the IBM simulation is a "black box" within the EF framework; as long as it can use inputs from the lifting function and produce outputs for the restriction function, any simulation method may be chosen.

System dynamics may be simulated by the lift-simulate-restrict cycle alone. 
However, it is computationally advantageous to use EF in conjunction with a projection step. In this approach, rather than simulating for an entire time-step, the simulation is run for a fraction of a time step to obtain an estimate the time derivative of the population-scale system. The population-scale state is then projected forward using a differential equation solver.

## Optimal control

@Clark1990 describes in detail the methods of deriving an optimal control path for dynamical ecological systems, which I briefly review here. The optimal path of a control variable over time $(h(t))$ is that which maximizes the integral of a profit function $\pi(x, h, t)$ over a time period $(T_0, T)$, subject to the dynamics of the system state $x$, the dynamics of which are governed by $\frac{dx}{dt} =  f(x, h, t)$.  To determine the optimal control path, one solves the maximization problem

$$\max_{h >= 0} \int_{T_0=0}^T \pi(x, h, t) \, dt\, 
  \text{ subject to } \frac{dx}{dt} =  f(x, h, t)$$

$x$ and $h$ may be vectors representing multiple system state and control 
variables.

Various numerical methods can be used to find the optimal control path. Where 
$f(x, t, h)$ has a closed-form, the control path can be derived analytically by 
maximizing the Hamiltonian ($\mathcal H$) equation:

$$\mathcal H = \pi(x, h, t) + \eta f(x, h, t)$$

Here $\eta$ is the "shadow value" of the system state $x$, and represents the per-unit value of the contribution $x$ to *future* profit.  The dynamics of $\eta$ are governed by the adjoint equation:

$$\frac{d\eta}{dt} = - \frac{\partial \mathcal H}{\partial x} = 
    -\frac{\partial \pi(\cdot)}{\partial x} + 
     \eta(t) \frac{\partial f(\cdot)}{\partial x}$$

To find the optimal path for the control $h(t)$, we solve for the maximum of $\mathcal H$ over the time period $(T_0, T)$.  The local optimum of $\mathcal H$ can be determined by finding $h$ where ${\partial \mathcal H}{\partial h} = 0$.

The resulting solution for $h(t)$ is dependent on the initial values of $x$ and $\eta$, $(x_0, \eta_0)$.  In many systems, some or all of these values are unknown, but terminal values ($x_T$, $\eta_T$) are constraints on the problem. In the case of unconstrained problems, the optimal path is that ends with $\eta_T = 0$. In these cases, initial conditions can be determined numerically using a boundary problem solver.

# Methods

## A framework equation-free optimal control

The method described above requires a closed-form expression for $dx/dt = f(x, h, t)$ in order to derive the optimal control path $h(t)$.  When the system of interest is modeled using an IBM, this population-level expression is not available. The EF framework can be used to calculate an optimal control path for an IBM, however, by numerically calculating $f(\cdot)$ and the values derived from it.  

In EF optimal control, $dx/dt, h(t)$, and $d\eta/dt$ are calculated numerically, and the optimal control path is calculated by using these values in a differential equation solver.  Starting at $t = T_0$, with initial conditions $x_0$ and $\eta_0$, $f(\cdot)$ is estimated numerically under an initial guess for $h(t)$ using a lifting-simulation-restriction cycle. Using this value of $f(\cdot)$, the value of $\mathcal H$ is calculated. This is repeated using a numerical optimizer to refine values of $h(t)$ until the value that maximizes $\mathcal H$, $h_{opt}(t)$,  is found.

To determine $d\eta / dt$, the value $\frac{\partial f(\cdot)}{\partial x}$ is required. This is estimated by perturbing $x$ by a small value, $\Delta x$, 
calculating $f(x + \Delta x, h_{opt}, t)$ and estimating $\frac{\partial f(\cdot)}{\partial x}$ by finite differencing. 

$dx/dt$ and $d\eta/dt$ can then be passed to a differential equation solver to 
project the forward system in time and calculate the paths of $x(t), \eta(t)$, and $h(t)$.  As in the optimal control scenario above, a boundary problem solver can then be used to determine initial values for $x_0$ and $\eta_0$ based on the constraints of the problem.

## Model system

I demonstrate the method described above on a problem of maintaining a wildlife population in the face of an invading disease. I model the system using an IBM closely related to the macroparasite model of @Anderson1978, where individuals 
can host multiple pathogens of the same type, and suffer increased mortality with greater infection load. The primary differences between @Anderson1978 and this model are the separation of birth and death processes and the influx of infectious particles from outside the system.

The control problem is to maximize net benefit of ecosystem services derived from the wildlife population by control measures reducing the influx of disease.  This is a problem faced in numerous systems, such as the conservation of frog populations in the face of *Batrachochytrium dendrobatidis* (chytrid fungus) [@Briggs2010], tanoak populations being invaded by *Phytophthora ramorum* (sudden oak death) [@Cobb2013], or bat populations at risk from *Geomyces destructans* (white nose syndrome) [@Langwig2015].  In each of these systems, invasion of the disease reduces local populations of hosts, but local populations can be protected from arrival of new disease with some efficacy by various measures, such as education, quarantine, disinfection, and culling nearby infected populations.

A deterministic, mean-field, continuous model of this system can be represented by a small set of ODEs. In this model, the host population $N$ increases via density-dependent reproduction ($rN(1-N/K)$) and decreases via a constant intrinsic mortality rate ($d$), as well as additional mortality $\alpha$ for each infection $P$ in a host:

$$\frac{dN}{dt} = r N (1 - N/K) - \alpha P - d N$$

The pathogen population $P$ grows via a spore production and establishment 
rate $lambda$ and via density-dependent contact between hosts, totaling $\lambda PN$. $P$ decreases via pathogen mortality ($\mu P$), background host mortality ($dP$), and mortality due to the disease, which affects the most-infected hosts most ($alpha (P + P/N)$). The form of this last term depends on the assumption of a random (Poisson) distribution of pathogens among hosts [@Anderson1978]. Finally, additional parasites enter the system via external propagule pressure $\lambda_{ex}$. $\lambda_{ex}$ may be reduced by control effort $h(t)$, which reduces propagule pressure by a factor of $e^{h(t)}$.

$$\frac{dP}{dt} = \lambda P N - \mu P - d P - \alpha P - \alpha P^2/N + e^{-h} 
N \lambda_{ex}$$

This model assumes (1) a large population size adequately represented by continuous variables, (2) a constant distribution of parasites among individuals, and (3) no stochastic processes.

The individual-based version of this model relaxes these assumptions. In the 
IBM, the state of the system is represented as a set of individual hosts, each with a discrete number of infections. Births, deaths, new infections and loss of infections (recovery) in each individual $i$ with number of infections $j_i$, occur stochastically according to rates $r$:

$$\begin{aligned}
r_{i, \, birth} &= \max \left\{ r(1-N/K), \, 0 \right\} \\
r_{i, \, death} &= \alpha j_i + d \\
r_{i, \, infection} &= \lambda \sum_{n=1}^N j_n + \lambda_{ex} e^{h(t)} \\
r_{i, \, recovery} &= \mu j_i
\end{aligned}$$

This stochastic process occurs in continuous time, implemented via Gillespie's [-@Gillespie1976] stochastic simulation algorithm (SSA).

Table 1 shows the model parameters used in all cases in this paper.

```{r loadpkgs, cache=FALSE, message=FALSE}
```

```{r setparms, cache=FALSE}
```

| Parameter | Symbol | Value |
|-----------|--------|-------|
| birth rate | $r$  | `r format(parms$r)` |
| carrying capacity | K | `r format(parms$K)` |
| intrinsic host mortality rate | $d$ | `r format(parms$d)` |
| intrinsic pathogen mortality rate | $\mu$ | `r format(parms$mu)` |
| additional mortality per infection | $\alpha$ | `r format(parms$alpha)` |
| contact rate | $\lambda$ | `r format(parms$lambda)` |
| external spore arrival rate | $\lambda_{ex}$ | `r format(parms$lambda_ex)` |
| initial host population | $N_0$ | `r format(macro_state[1])` |
| initial pathogen population | $P_0$ | `r format(macro_state[2])` |
| benefit per host per unit time | $v$ | `r format(parms$v)` |
| cost of control per unit time  | $c$ | `r format(parms$c)` |
| time period | $T$ | `r format(parms$time_max)` |

  Table: Parameters for the host-pathogen models and management problem

## Management problem

For the management problem, $v$ is the value of ecosystem service provided per host per unit time. The control variable, $h$, represents of the effort expended per unit time in reducing the arrival rate of new infections and each unit of effort of control, and cost $c$. The optimization problem is to maximize the net benefits over the course of a fixed time period $T$, subject to the dynamics of the system.

$$\max_{h >= 0} \int_{t=0}^T  \pi(x,h,t), dt\, \text{ s.t. } \frac{dx}{dt} = 
f(x, h, t) $$

I do not include discounting in this problem definition.

For the mean-field ODE system, one can derive the Hamiltonian equation based on 
this problem, which must be optimized for all time points, the shadow values 
$\eta_1$ and $\eta_2$, and an expression that can be solved for $h(t)$:

$$\begin{aligned}
\mathcal{H} &= vN - ch + \\
&\eta_1 \left( rN(1 - N/K) - \alpha P - d N \right) + \\
&\eta_2 \left(\lambda P N - \mu P - d P - \alpha P - \alpha (P^2)/N + e^{-h}  N 
\lambda_{ex} \right) \\
\frac{d\eta_1}{dt} &= -v - \eta_1 \left(r - d - 2r \frac{N}{K}\right) - \eta_2 
\left(\lambda P + \alpha \frac{P^2}{N^2} + e^{-h} \lambda_{ex} \right) \\
\frac{d\eta_2}{dt} &= \eta_1 \alpha - \eta_2 \left(\lambda N - \mu - d - \alpha - 
\frac{2 \alpha P}{N} \right) \\
0 &= h \left( -c - \eta_2 e^{-h} N * \lambda_{ex} \right)
\end{aligned}$$

# Equation-free numerical approach

To solve the management problem under the IBM system, I use the equation-free approach described above. As in this case, the Hamiltonian can only be expressed in terms of the estimated values for $f(\cdot)$.  I use dot notation $(\dot P \approx dP/dt)$ to denote derivatives estimated the EF lift-simulate-restrict cycle:

$$\begin{aligned}
\bar{\mathcal{H}} &= vN - ch + \eta_1 \dot N + \eta_2 \dot P \\
\dot \eta_1 &= -v - \eta_1 \frac{d \dot N}{dN} - \eta_2 \frac{d \dot P}{dN} \\
\dot \eta_2 &= -v - \eta_1 \frac{d \dot N}{dP} - \eta_2 \frac{d \dot P}{dP}
\end{aligned}$$

I use a lifting function that randomly distributes $P$ infections across $N$ individuals (a Poisson process). For the reverse (restriction), I simply sum the total host and parasite populations. When mapping from non-integer population-scale states to individual-scale states, simulations are run with all combinations of macro-variables rounded up and down, and the overall results is determined by a weighted average of these simulations.

To determine the estimated control value at each time point $(\bar h_{opt})$, I 
use a numerical maximization routine [BOBYQA, @Powell2009a], to find the value 
of $h$ that maximizes the estimated Hamiltonian $\bar{\mathcal{H}}$.

While the EF framework can be used on "black box" simulators, taking advantage of some properties of the individual-based model can improve computational efficiency and performance of the method. In this case, I do so in two ways. First, as the IBM uses the Gillespie simulation algorithm, I use a single variable-time Gillespie IBM rather than a fixed-period. This allows me to estimate $\dot N$ and $\dot P$ directly at each time step $t$, avoiding the error incurred with a fixed step which would estimate these values at a time slightly offset from $t$. Second, rather than rather than estimating $d \dot X /dX$ terms above by perturbing the system after solving for $\bar h_{opt}$, I take advantage of the mapping between continuous population states at the macro-level and discrete counts at the micro level. When lifting a continuous population-scale state to create an ensemble of random discrete IBM states, some have populations rounded down from the continuous level, and some rounded up. The difference in $\dot X$ between these instances of the simulation is used to determine $d \dot X/ dX$ without additional simulations.

The system is integrated using a simple forward Euler integrator and then initial conditions for boundary problem solved via a shooting algorithm. More sophisticated and computationally intense integrators, such as Runge-Kutta and Adams-Bashford may be used for greater accuracy [@Williams2015].

As both the maximization routine and the forward integrator are quite sensitive 
to noise in the estimated derivatives, solving the system required large ensembles of IBM simulations in the EF cycle. To calculate $\dot h_{opt}$, I used 4 million simulations, and on the final iteration in each time step, 400 million simulations to calculate the values of $d\eta/dt$. The entire algorithm described above took approximately 12 hours to solve on a computer with 24 2.2GHz Intel Processors. code for the simulation is archived on GitHub and Zenodo. 
[TODO:ARCHIVE AND REFERENCE CODE].

# Results


```{r ODEsim, include=FALSE}
```
```{r IBMsim, include=FALSE}
```
```{r EFsim, include=FALSE}
```

Figure 1 shows trajectories for the system in the absence of control, simulated 
in three ways: the mean-field ODE system, `r format(parms$n_comp_sims)` simulations of the IBM, and the expected trajectory as estimated by the equation-free model with `r format(parms$n_sims)` simulations per step.

```{r f1, include=TRUE, echo=FALSE, cache=FALSE, fig.height = 6, fig.width = 11, fig.cap="Trajectories of the host-pathogen system without control. In cases host population shown in blue and pathogens shown in red. (A) The system simulated using the ODE model (solid lines) (B), The system as simulated via 100 instances of the IBM model. Individual runs are in light colors. Mean values at each time point shown with dashed lines, dotted lines at ± 2 standard deviations, (C) The system simulated with the EF method (dash-dotted lines). (D) Trajectories from all three methods overlain for comparison. ODE (solid) and EF (dot-dash) lines overlap exactly. Mean of IBM runs shown in dashed lines."}
<<fig1>>
FIG1
```

The general trajectory of the system is similar in all cases. In the absence 
of control, the disease invades the system rapidly. This reduces the population 
via increased overall mortality rate for hosts, and the system reaches a stable equilibrium with a suppressed host population endemic infections.

While the trajectories of all three models are similar, the EF value is closer to the mean-field ODE than the average of the IBM trajectories. This reflects the 
fact that both the analytical derivation of the ODE system from @Anderson1978 and the choice of lift/restrict functions in the EF model make similar assumptions. Both assume that the distribution of infections follows a Poisson distribution.  
This difference of the complete IBM reflects some deviation from the Poisson assumption.

```{r ODEsoln, include=FALSE, cache=TRUE}
```

Figure 2 shows the optimal control path as calculated by the analytical method on the ODE system. Under this set of parameters, the optimal control path for the system is one of declining effort and abandonment. Initially, high control effort reduces the rate of increase of the pathogen, resulting in higher host populations over the early management period, which provides greater benefits. The optimal effort path then declines, allowing greater pathogen growth, until the pathogen population reaches a peak, at which point the optimal strategy is abandonment. At this late stage, external spore arrival rate has little effect; the pathogen population is limited by internal factors. Shadow values show that the value of the host population declines over the course of the management period, and the negative value of the pathogen increases. Later in the period, the pathogen has less potential to reduce overall profits.  

```{r f2, include=TRUE, echo=FALSE, cache=FALSE, fig.cap="Trajectory of the ODE system under optimal control.  (A) Trajectories of host and pathogen populations under control. (B) Trajectories of host and pathogen shadow values. (C) Trajectory of control effort over time."}
<<fig2>>
FIG2
```

Figure 3 shows the paths for the solution as calculated using the EF
method. The control strategy, as well as host and pathogen population dynamics, is essentially the same as in the ODE approach. Small numerical errors in the EF method result in imprecision in the shooting algorithm, resulting in terminal shadow values slightly off from zero. In this case, these imprecisions have little effect on the result of the algorithm as they accumulate towards the end of the control period, after point when the optimal control effort falls to zero. Figure 4 shows that the control path from the EF method yields similar profit values to the ODE-derived control path.

```{r f3, include=TRUE, echo=FALSE, cache=FALSE, fig.cap="Trajectory of the EF system under optimal control. (A) Trajectories of host and pathogen populations under control. (B) Trajectories of host and pathogen shadow values. (C) Trajectory of control effort over time."}
<<fig3>>
FIG3
```

```{r f4, include=TRUE, echo=FALSE, cache=FALSE, fig.cap="Net profit under no control and optimal control for both the ODE system and the equation-free (EF) system."}
<<fig4>>
FIG4
```

# Discussion and Conclusions

I have shown that a numerical approach using the EF framework can recover the optimal control path of an IBM. In this test case, the system could be reduced to a tractable set of analytical equations with the assumption of Poisson-distributed infections among individuals. Using the same assumption under the EF approach, I was able to derive an identical control path as the analytical system purely numerically, with the same net value to the manager. The EF approach should perform more accurately than analytical approaches when it makes less drastic assumptions. For instance, in an alternate formulation of the @Anderson1978 macroparasite infections are aggregated and represented by a negative-binomial distribution with aggregating parameter $k$, but one assumption made is that $k$ is constant, though this has been shown not to hold  over the course of an epidemic [@Adler1992, and Chapter 3 of this dissertation]. An EF representation of aggregated infections could include $k$ as a state variable and provide more accurate simulation and solution to economic problems. A variety of other individual-based and structured population models useful for disease management have high dimensionality and are not easily reduced, such as disease spread in spatial point processes [@Brown2004], and through networks [@Gross2008a; @Reppas2010], and agent-based models where individuals exhibit economic behavior [@Fenichel2010].

One important area for exploration is understanding the performance of the EF approach on stochastic systems. EF modeling has been shown to effectively capture important properties of stochastic population models such as extinction probability and extinction time [@Williams2015], and can provide the expected value of stochastic system under conditions where they might differ from deterministic reductions, as in the case of small populations. However, optimal economic strategies derived from the expected behavior of stochastic systems may not be optimal across all possible stochastic outcomes.

The method described here is computationally intense, primarily due to the large number of simulations required to estimate derivatives and partial derivatives at each time step to sufficient accuracy. A number of approaches may potentially improve its speed, accuracy, and robustness. More sophisticated forward integrators, such as adaptive Adams-Bashforth methods can be used [@Williams2015] for results that are more accurate. Bayesian optimization [@Snoek2012] may be more efficient in maximizing a Hamiltonian equation with noisy estimates for system derivatives. The IBM used here, as well as others using the stochastic simulation algorithm, could be sped up using $tau$-leaping [@Gillespie2001] or similar improvements. Also, the EF approach may be applicable to a different general method of continuous optimal control, piecewise polynomial curve-fitting [@Sirisena1973], which is more robust than shooting-based boundary value problem solvers to bifurcations and discontinuities.

This equation-free approach is best suited to cases where control effort may change continuously over time. Depending on the system and management problem, other optimization techniques that can operate on "black-box" IBMs may be used. Reinforcement learning [@Sutton1998], may be used for optimization using a black-box simulation model in cases with discrete control periods and discrete sets of control choices. Like EF modeling, reinforcement learning requires the careful selection of population-level variables summarizing the state of an underlying IBM which are relevant to the control problem and IBM dynamics.

Modeling disease in systems with population structure, individual variation, and stochasticity or combinations thereof often requires IBMs for which governing equations are unknown or intractable. This method provides a new tool for solving economic problems where such complex models are required. The EF framework holds promise for solving ecological, theoretical, and management problems with IBMs.

# References

