---
title: "Optimal Control of Individual-Based Disease Models"
author: "Noam Ross"
date: "July 25, 2015"
output: html_document
bibliography: references.bib
csl: ecology-letters.csl
---


```{r get_chunks, echo=FALSE, results='hide', message=FALSE, warning=FALSE, cache=FALSE}
library(knitr)
read_chunk('base_parms.R')
read_chunk('compare_graphs.R')
opts_knit$set(eval.after = "fig.cap")
opts_chunk$set(echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE)
```

## Introduction

A central challenge in disease ecology is developing effective management 
strategies to control harmful diseases and limit their damage to wildlife, 
human health, and ecosystem services. Using models of the dynamics of hosts and 
disease, this can be formulated as an economic optimization problem: how to 
maximize the benefits of a strategy, or how to minimize the costs of meeting a 
disease control goal.

Disease management may be considered an economic problem: Given limited 
resources, how to best manage disease to minimize impacts? Or, conversely, 
given management targets, how to best limit the cost of disease control? 
Research addressing such questions - economic epidemiology - is a growing field 
[@Brandeau2003; @Rowthorn2009], though the literature to on the economics of 
wildlife disease is more limited] [Bicknell1999; @Horan2005; Fenichel2010; 
@Horie2013]. The bioeconomics of invasive species management is more broadly 
explored [@Sharov1998; @Carrasco2010; @Haight2010; @EpanchinNiell2012], and has 
many close parallels with the economics of disease management 

Individual-based models (IBMs, also called agent-based models) simulate systems 
based on the behavior and properties of individuals, rather than the population 
as a whole. IBMs can simulate system behavior arising from interactions 
difficult to describe in aggregate, including network contact structure, 
spatial heterogeneity, variation in individual traits, all of which can affect 
disease transmission. These individual behaviors and interactions in such 
models result in emergent population-level dynamics that often can not be 
expressed as a closed-form set of equations at the population level without 
making strong assumptions about the distribution of individuals. [@Grimm2005]

IBMs can and have been used to study disease economics in a number of ways.  
The first, and most common, is to simulate IBMs of disease systems under a 
number of discrete management scenarios and examine the outcomes of each 
scenario to determine which does best according to performance criteria [TODO:REF]. 
 Secondly, IBMs can incorporate adaptive or optimizing behavior within 
individuals in order to better simulate disease dynamics [@Fenichel2010]. 

Determining optimal control approaches using IBMs is more challenging. In the 
absence of closed-form equations of overall system behavior, one can not derive 
optimal control decisions analytically.  IBMs also suffer from "the curse of 
dimensionality": high dimensionality results in very large state space that 
can not be explored thoroughly.
 
One approach to optimal controlling using IBMs has been to apply control 
analytical techniques to reduced-form models. The optimization problem can be 
solved on a set of ordinary differential equations that represent an 
approximation of the IBM. Using this approach, a mean-field approximation of an 
IBM can be used to determine an optimal control path when the IBM is 
homogenous, but ay perform poorly under when the underlying system is 
heterogenous [@Federico2013].  Another approach is to generate population-level 
models with non-mechanistic equations that approximate IBM behavior. Such 
methods can yield better results than mean-field models. However, the 
applicability of these adjusted models is limited to the the range over which 
they were parameterized, and model parameters are difficult to interpret 
[Oremland 2015].

Here I describe another approach to deriving optimal control strategies for 
ecological systems using IBMs that uses equation-free, or multi-scale, modeling 
[Armaou2004] to estimate mean dynamics of an IBM numerically and estimate 
quantities that allow the use of analytic tools to determine an optimal control 
path.  First, I introduce the equation-free modeling framework.  I then 
describe a disease system and equivalent ODE and IBM models.  I solve the ODE 
system using classic tools and the IBM model using the methods described, 
showing that they generate the same solution.

## Equation-free modeling

Equation-free (EF), or multi-scale, modeling (@Armaou2004, but see 
@Kevrekidis2009a for review) is a method for capturing and analyzing 
population-scale dynamics of individual-based models while bypassing the 
deriving population-scale equations, which exist conceptually but often can not 
be derived. It has been used in a variety of applications, including evolution, 
movement ecology and epidemiology [@Cisternas2004; [@Erban2006; Gross2008a; 
Raghib2010; @Williams2015]

In EF modeling, the population-scale dynamics are estimated using bursts of the 
IBM. At each time step, a "lifting" operator maps the population-scale state to 
a distribution of possible individual-scale states. Each of these instances of 
the individual-based model is simulated for a short burst of time, then they 
are collapsed back to a single population-level state with a "restricting 
function."  

Some information about the IBM state is lost by restriction and generated by 
lifting.  Thus, these EF is another form of model reduction, albeit one that 
offers alternatives to mean-field and other approximations.  Lifting and 
restriction operators may be chosen to capture the important heterogeneities 
represented in the IBM and relevant to the problem at hand.  While these 
functions must be chosen with care, the rules governing the IBM may be a "black 
box" within the EF framework.

The system dynamics may be simulated by the lift-simulate-restrict cycle alone. 
However, it is computationally advantageous to use in conjunction with a 
projection step. In this approach, rather than simulating for an entire 
time-step, the simulation is run for a fraction of a time step to obtain an 
estimate the time derivative of the population-scale system. The 
population-scale state is then projected forward using a differential equation 
solver. 

## Optimal control

@Clark1990 describes in detail the methods of deriving an optimal control path 
for dynamical ecological systems, which I briefly review here. The optimal path 
of a control variable over time $(h(t))$ is that which maximizes the integral 
of a profit function $\pi(x, h, t)$ over a time period $(T_0, T)$, subject to 
the dynamics of the system state $x$, the dynamics of which are governed by 
$\frac{dx}{dt} =  f(x, h, t)$.  That is,

$$\max_{h >= 0} \int_{T_0=0}^T \pi(x, h, t) \, dt\, 
  \text{ subject to } \frac{dx}{dt} =  f(x, h, t)$$.



Where $x$ and $h$ may be vectors representing the system state and control 
variables.

Various numerical methods can be used to find the optimal control path. Where 
$f(x, t, h)$ has a closed-form, the control path can be derived analytically by 
maximizing *Hamiltonian* ($\mathcal H$) equation:

$$\mathcal H = \pi(x, h, t) + \eta f(x, h, t)$$.

Here $\eta$ is the "shadow value" of the system state $x$, and represents the 
per-unit value of the contribution $x$ to *future* profit.  The dynamics of 
$\eta$ are governed by the adjoint equation: 

$$\frac{d\eta}{dt} = - \frac{\partial \mathcal H}{\partial dx} = 
    -\frac{\partial \pi(\cdot)}{\partial x} + 
     \eta(t) \frac{\partial f(\cdot)}{\partial x}$$

To find the optimal path for the control $h$, we solve for the maximum of 
$\mathcal H$ over the time period $(T_0, T)$.  The local optimum of $\mathcal 
H$ can be determined by finding $h$ where ${\partial \mathcal H}{\partial dh} = 
0$.

The resulting solution for $h(t)$ is dependent on the initial values of of $x$ 
and $\eta$, $(x_0, \eta_0)$.  In many systems, some or all of these values are 
unknown.  In many cases the terminal values ($x_T$) are constraints on the 
problem. In the case of unconstrained problems, the optimal path is that that 
ends with $\eta_T = 0$.  In these cases, initial conditions can be determined 
numerically using a boundary problem solver.

## A framework equation-free optimal control

The method described above requires a closed-form expression of $dx/ft = f(x, 
h, t)$ in order to derive the optimal control path $h(t)$.  When the system of 
interest is modeled using an IBM, this population-level expression may not be 
available, even while our objective function $\pi(\cdot)$ is expressed in terms 
of population-level state.   The EF framework can be used to calculate an 
optimal control path for an IBM, however, by numerically estimating $f(\cdot)$ 
and the values derived from it.  

In EF optimal control, $f(\cdot) = dx/dt, d\eta/dt$, and $h_{opt}(t)$, are 
estimated numerically at each time step in a differential equation solver.  
Starting at $t = T_0$, with initial conditions $x_0$ and $\eta_0$, $f(\cdot)$ 
is estimated numerically under an initial guess for $h(t)$ using a short burst 
of the lifting-simulation-restriction cycle.  Using this value of $f(\cdot)$, 
the value of $\mathcal H$ is calculated.  Using a numerical optimizer, this is 
repeated with refined values of $h(t)$ until a value that maximizes $\mathcal 
H$,  $h_{opt}(t)$ is found.

To determine $d\eta / dt$, an value for $\frac{\partial f(\cdot)}{\partial x}$ 
is required.  This is estimated by perturbing $x$ by a small value, $\Delta x$, 
calculating $f(x + \Delta x, h_{opt}, t)$ and estimating $\frac{\partial 
f(\cdot)}{\partial x}$ by finite differencing.  

$dx/dt$ and $d\eta/dt$ can then be passed to a differential equation solver to 
project the forward system in time and calculate the paths of $x(t), \eta(t)%, 
and $h(t)$.  As in the optimal control scenario above, a boundary problem 
solver can then be used to determine initial values for $x_0$ and $\eta_0$ 
based on the constraints of the problem.

## Model system

I demonstrate the method described above on a problem of maintaining a wildlife 
population in the face of an invading disease.  The model is a host-pathogen 
system similar to the macroparasite model of @Anderson1978, where individuals 
can host multiple pathogens of the same type, and suffer increased mortality 
with greater infection load. The primary differences are the separation of 
birth and death processes and the influx of infectious particles from outside 
the system.

The control problem is to maximize net benefit of ecosystem services derived 
from the wildlife population by costly control measures reducing the influx of 
disease.  This is a problem faced in numerous systems, such as the conservation 
of frog populations in the face of *Batrachochytrium dendrobatidis* (chytrid 
fungus), tanoak populations being invaded by *Phytophthora ramorum* (sudden oak 
death), or bat populations at risk from *Geomyces destructans* (white nose 
syndrome).  In each of these systems, invasion of the disease reduces local 
populations of hosts, but local populations can be protected from arrival of 
new disease with some efficacy by various measures, such as education, 
quarantine, disinfection, and culling nearby infected populations.

A deterministic, mean-field, continuous model of this system can be represented 
by a small set od ODEs.  In this model, the host population $N$ increases via 
density-dependent reproduction ($rN(1-N/K)$) and decreases via a constant 
intrinsic mortality rate ($d$), as well as additional mortality $\alpha$ for 
each infection $P$ in a host:

$$\frac{dN}{dt} = r N (1 - N/K) - \alpha P - d N$$

The pathogen population $P$ grows via a spore production and estabilishment 
rate $lambda$  and via density-dependent contact between hosts, totaling 
$\lambda P N$.  $P$ decreases via pathogen mortality ($\mu P$), background host 
mortality ($dP$), and mortality due to the disease, which affects the 
most-infected hosts most $alpha (P + P/N)$. The form of this term depends on 
the assumption of a random (Poisson) distribution of pathogens among hosts 
[@Anderson1978]. Finally, additional parasites enter the system via external 
propagule pressure $\lambda_{ex}$. $\lambda_{ex}$ may be reduced by control 
effort $h(t)$, which reduces propagule pressure by a factor of $e^{h(t)}$.

$$\frac{dP}{dt} = \lambda P N - \mu P - d P - \alpha P - \alpha P^2/N + e^{-h} 
N \lambda_{ex}$$

This model assumes (1) a large population size adequately represented by 
continuous variables, (2) a constant distribution of parasites among 
individuals, and (3) no stochastic processes.

The individual-based version of this model relaxes these assumptions. In the 
IBM, the state of the system is represented as a set of individual hosts, each 
with a discrete number of infections. Births, deaths, new infections and loss 
of infections (recovery) in each individual $i$ with number of infections 
$j_i$, occur stochastically according to rates $r$:

$$\begin{aligned}
r_{i, \, birth} &= \max \left\{ r(1-N/K), \, 0 \right\} \\
r_{i, \, death} &= \alpha j_i + d \\
r_{i, \, infection} &= \lambda \sum_{n=1}^N j_n + \lambda_{ex} e^{h(t)} \\
r_{i, \, recovery} &= \mu j_i
\end{aligned}$$

This stochastic process occurs in continuous time, implemented via Gillespie's 
[-@Gillespie1976] stochastic simulation algorithm (SSA).

Table 1 shows the model parameters used for all models in this paper.

```{r loadpkgs}
```

```{r setparms}
```

| Parameter | Symbol | Value |
|-----------|--------|-------|
| birth rate | $r$  | `r format(parms$r)` |
| carrying capacity | K | `r format(parms$K)` |
| intrinsic host mortality rate | $d$ | `r format(parms$d)` |
| intrinsic pathogen mortality rate | $\mu | `r format(parms$mu)` |
| additional mortality per infection | $\alpha$ | `r format(parms$alpha)` |
| contact rate | $\lambda$ | `r format(parms$lambda)` |
| external spore arrival rate | $\lambda_{ex}$ | `r format(parms$lambda_ex)` |
| initial host population | $N_0$ | `r format(macro_state[1])` |
| initial pathogen population | $P_0$ | `r format(macro_state[2])` |
| benefit per host per unit time | $v$ | `r format(parms$v)` |
| cost of control per unit time  | $c$ | `r format(parms$c)` |
| time period | $T$ | `r format(parms$time_max)` |

  Table: Parameters for the host-pathogen models and management problem

### Management problem

For the management problem, the objective function is to is to maximize 
ecosystem service provided by the host population, $v$, per host per unit time. 
The control variable, $h$, represents of reducing the arrival rate of new 
infections and each unit of effort has a cost $c$. Thus, the optimization 
problem is to maximize the net benefits over the course of a time period $T$, 
subject to the dynamics of the system.

$$\max_{h >= 0} \int_{t=0}^T  \pi(x,h,t), dt\, \text{ s.t. } \frac{dx}{xt} = 
f(x, h, t) $$

For the mean-field ODE system, one can derive the Hamiltonian equation based on 
this problem, which must be optimized for all time points, the shadow values 
$\nu_1$ and $\nu_2$, and an expression that can be solved for $h(t)$:

$$\begin{align}
\mathcal{H} &= vN - ch + \\
&\nu_1 \left( rN(1 - N/K) - \alpha P - d N \right) + \\
&\nu_2 \left(\lambda P N - \mu P - d P - \alpha P - \alpha (P^2)/N + e^{-h}  N 
\lambda_{ex} \right)

\frac{d\nu_1}{dt} &= -v - \nu_1 \left(r - d - 2r \frac{N}{K}\right) - \nu_2 
\left(\lambda P + \alpha \frac{P^2}{N^2} + e^{-h} \lambda_{ex} \right) + \delta 
\nu_1 \\

\frac{d\nu_2}{dt} &= \nu_1 \alpha - \nu_2 \left(\lambda N - \mu - d - \alpha - 
\frac{2 \alpha P}{N} \right) + \delta \nu_2 \\

0 &= h \left( -c - \nu_2 e^{-h} N * \lambda_{ex} \right)
\end{align}$$

### Numerical approach

To solve the management problem under the IBM system, I use the equation-free 
approach described above.  As in this case, the Hamiltonian
can only be expressed in terms of the estimated values for $f(\cdot)$.  I use 
dot notation $(\dot P \approx dP/dt)$ to denote derivatives estimated the EF 
lift-simulate-restrict cycle:

$$\begin{align}
\bar{\mathcal{H}} &= vN - ch + \nu_1 \dot N + \nu_2 \dot P \\
\dot \nu_1 &= -v - \nu_1 \frac{d \dot N}{dN} - \nu_2 \frac{d \dot P}{dN} \\
\dot \nu_2 &= -v - \nu_1 \frac{d \dot N}{dP} - \nu_2 \frac{d \dot P}{dP}
\end{align}$$

I use a lifting function that’s randomly distributes $P$ infections across $N$ 
individuals (a Poisson Process).  For the reverse (restriction), I simply sum 
the total host and parasite populations.  When mapping from non-integer 
macro-states, simulations are run with all combinations of macro-variables 
rounded up and down, and the overall results is determined by a weighted 
average of these simulations.

To determine the estimated control value at each time point $(\bar h_{opt})$, I 
use a numerical maximization routine (BOBYQA, @Powell2009), to find the value 
of $h$ that maximizes the estimated Hamiltonian $\bar{\mathcal{H}}$.

While the EF framework can be used on "black box" simulators, taking advantage 
of some properties of the individual-based model can improve computational 
efficiency and performance of the method. In this case, I do so in two ways. 
First, as the IBM uses the Gillespie simulation algorithm, I use a single 
variable-time "Gillespie" micro-step rather than a fixed-period micro-step. 
This allows me to estimate $\dot N$ and $\dot P$ directly at each time step 
$t$, avoiding the error incurred with a fixed micro-step which would estimate 
these values at a time slightly offset from $t$. Second, rather than rather 
than estimating $d \dot X /dX$ terms above by perturbing the system after 
solving for $\bar h_{opt}$, I take advantage of the mapping between continuous 
population states at the macro-level and discrete counts at the micro-level. 
When lifting a continuous macro-level to create an ensemble of random 
micro-level states, some will have populations rounded down from the continuous 
level, and some rounded up. The difference in $\dot X$ between these instances 
of the simulation is used to determine $d \dot X/ dX$ without additional 
simulations.

The system is integrated using a simple forward Euler integrator and then a 
boundary problem solved via a shooting algorithm.  More sophisticated and 
computationally intense integrators, such as Runge-Kutta and Adams-Bashford may 
be used for greater accuracy [@Williams2015].

As both the maximization routine and the forward itegrator are quite sensitive 
to noise in the estimated derivatives, solving the system required large 
ensembles of IBM simulations in the EF cycle.  To calculate $\dot h_{opt}$, I 
used `r format(parms$n_sims)` simulations, and on the the final iteration in each time 
step, `r format(parms$n_sims_jacob)` simulations. The entire algorithm described above took 
approximately 8 hours to solve on a computer with 24 2.2GHz Intel Processors.  Code for the simulation is archived on GitHub and Zenodo. 
[TODO:ARCHIVE AND REFERENCE CODE].

Results
-------

```{r ODEsim, include=FALSE}
```
```{r IBMsim, include=FALSE}
```
```{r EFsim, include=FALSE}
```

Figure 1 shows trajectories for the system in the absence of control, simulated 
in three ways: the mean-field ODE system, `r format(parms$n_comp_sims)` simulations 
of the IBM, and the expected trajectory as estimated by the equation-free model 
with `r format(parms$n_sims)` simulations per step.

The general trajectory of the system is similar in all cases.  In the absence 
of control, the disease invades the system rapidly. This reduces the population 
via increased overall mortality rate for hosts, after a peak in the pathogen 
population, the system reaches a stable equilibrium with a suppressed host 
population.

While the trajectory of all three models is similar, the EF value is closer to 
the mean-field ODE than the average of the IBM trajectories.  The match between 
the EF and the mean-field ODE model confirms that the Poisson lifting function 
is equivalent to the Poisson assumption in the set of ODEs, making the two 
models functionally equivalent. The deviation of the mean of many IBMs from the 
EF and ODE models can be attributed to Jensen's [-@Jensen1906] inequality.

Figure 2 shows the optimal control path 

In the ODE system, the internal equilibrium can be calculate as [TODO].

The individual-based, stochastic model results in figure. Shown are the mean 
path and the 95% confidence intervals for the system state through time,

-   IBM and ODE versions of the model have different behavior
-   EF version of IBM correctly produces results similar to overall IBM:
    -   Plots: Dynamics of ODE, IBM (average of many runs), and EF versions of 
model under conditions of similar behavior (large population, equilibrium), and 
different behavior (small population, transients)
-   IBM and ODE have different responses to the the same controls.
    -   Characteristics of the control path as solved for the ODE
        -   Plot: ODE dynamics under control
        -   Plot: IBM dynamics (average and several example runs) under 
ODE-derived control
-   EF optimization yields better results for the IBM than ODE-derived control
    -   Plot: IBM dynamics under EF optimization
    -   Plot: Compare net present value of
        -   IBM and ODE models without control
        -   IBM and ODE models with ODE-derived control
        -   IBM under EF-derived control.

Discussion
----------

### Implications for disease control

-   [Short section, this is really a separate, less method-intense paper]
-   External control only has limitations; best used to reduce impacts, not 
prevent outbreak
-   Quantitative implications of stochasticity

### Methodological issues

This IBM is relatively simple. It lacks several features that are trivial to 
incorporate into an IBM but challenging to represent in closed-form solutions, 
such as individual variation in traits or heterogenity in the host contact 
structure.


-   Importance of lifting, restriction functions
-   Sensitivity to number of simulations and micro-step size, integration 
algorithms
-   Boundary conditions. ODE can provide guidance in values.

### Broad implications

-   EF-based optimization has applications on a wide variety of models and 
situations.
-   Complement to discrete approaches, ensure agreement with theory.

-   Still a model reduction approach, but reduction takes place at a different 
scale
-   Advantages
    -   Continuous in time and space.
    -   Interpretability. Model-fitting approaches to reduced models leave us 
with uninterpretable values.
    -   Also applies to black-box processes
-   Challenges
    -   Computational cost (though many methods have this, and Hamiltonian 
approach avoids much of it)
    -   Lifting and restriction functions must be chosen carefully. Model 
reduction occurring at the micro-scale.
    -   Software
        -   Some details of R package
-   Stochasticity: The EF model produces an estimate of the IBM's *mean* 
behavior, and thus an optimal control path for the mean conditions. The primary 
utility of the method is not in stochastic control. [See a host of other 
methods for this: Talk to Carl/Jim]. Rather, its value lies in optimizing where 
the IBM behaviors are not easily expressed in closed form equations. Stochastic 
behavior makes the method computationally difficult. The variation in outcomes 
is amplified in the estimate of the derivatives and even more so in the 
estimation of of the Jacobian matrix. \#\#\# Future work

